---
title: 'Course Project: Shiny App Pitch'
author: "Paul Springer"
date: "5th of August 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## General App Description

This App allows the user to predict tooth length of a pig depending on the dosage
of vitamin C and on the delivery method of vitamin C. Two delivery methoda are
considered: OJ - orange juice and VC - ascorbin acid. The data are taken from
standard R data set ToothGrowth.

## Available Regression Models

The User can decide between linear or quadratic prediction models. Model choice strongly
influences the predictions. Here is an example for OJ and dosage = 1:

```{r}
data("ToothGrowth")
```


```{r, echo = TRUE}
modelOJ_lin <- lm(len ~ dose, data = ToothGrowth[ToothGrowth$supp ==
                                                         "OJ", ])
modelOJ_quad <- lm(len ~ dose + I(dose^2), data =
                           ToothGrowth[ToothGrowth$supp == "OJ", ])
predict(modelOJ_lin, newdata = data.frame(dose = 1))
predict(modelOJ_quad, newdata = data.frame(dose = 1))
```


## Data used

```{r, echo = TRUE}
head(ToothGrowth)
str(ToothGrowth)
```

## Resources

+ The Application can be found under following link:

https://paulspringer.shinyapps.io/projectapp/

+ Documentation of the application is included in application itself in tab 
"Documentation".

+ The source code (ui.R and server.R) and this presentation can be found under:

https://github.com/PaulSpringer/datasciencecoursera
